---
title: "MSCI718_Final_Project"
author: "Aayushi Beniwal"
date: "08/04/2022"
output: html_document
---
## Section 1: R Programming Question 1
### 1 (a) 
Simulating the process once by generating the arrival time of patients using exp function as it is
exponentially distributed. After that I am calculating the waiting time for the patient by calculating the index value of the min doc appoint time(out if 3 doctors) and then on that index value updating the new waiting time for the doc by summing up prev arrival time with uniformly distributed number using runif (between 5 mins to 20 mins) which gives one random value. By taking  mean of the waiting time I am calculating final avg mean time. The max time left in the doctor appoint time list will be the closing time for the hospital.
```{r}
# Section 1: R Programming Question 1

# ________ Question 1 (a) _______
#(a) Simulating the process once. 
set.seed(7)
# 9 am to 4 pm will be 420 mins
max_time <- 420
last_time <- 0
waiting_time <- c()
tot_waiting_patients <- 0
tot_patients <- 0
# there are 3 doctors in clinic
doc_appoint_time <- c(0, 0, 0)

while (last_time < max_time) {
  # arrival time of patients is exponentially distributed with expectation of 10 mins
  time_arrival <- last_time + rexp(1, rate = 1/10)
  tot_patients <- tot_patients + 1
  last_time <- time_arrival
  
  min_doc_appoint_time <- which.min(doc_appoint_time)
  wait_time <- max(c(0, doc_appoint_time[min_doc_appoint_time] - time_arrival))
  if (wait_time > 0) {
    # Patients waiting for a doctor
    tot_waiting_patients <- tot_waiting_patients + 1
    waiting_time <- append(waiting_time, wait_time)
  }
  # time spent by doc on each patient is uniform between 5 to 20 mins
  doc_appoint_time[min_doc_appoint_time] <- time_arrival + wait_time + runif(1, 5, 20)
}
# avg wait time for patients 
if (length(waiting_time) == 0) {
  avg_waittime <- 0
}else {
  avg_waittime <- mean(waiting_time)
}

# Office closing time
office_close <- max(doc_appoint_time)
office_close_time <- office_close - max_time

# summary of outputs
df1a <- data.frame(tot_patients = tot_patients,tot_waiting_patients = tot_waiting_patients,
                   avg_waittime = avg_waittime)
df1a
print(paste(" Office close time: 4:",as.integer(office_close_time),"p.m."))
```

### 1 (b) 
Simulating the above process 1000 times and saving the final values in a list and append in it every time so that after the for loop I can take the final median values.
```{r}
# ________ Question 1 (b) _______
#(b) Simulating the process 1000 times 
set.seed(7)
num_simulations <- 1000
avg_waiting_time_l <- c()
office_close_time_l <- c()
tot_patients_list <- c()
tot_patients_waiting_l <- c()


for (i in 1:num_simulations) {
  last_time <- 0
  waiting_time <- c()
  tot_waiting_patients <- 0
  tot_patients <- 0
  # there are 3 doctors in clinic
  doc_appoint_time <- c(0, 0, 0)
  while (last_time < max_time) {
    # arrival time of patients is exponentially distributed with expectation of 10 mins
    time_arrival <- last_time + rexp(1, rate = 1/10)
    tot_patients <- tot_patients + 1
    last_time <- time_arrival
    
    min_doc_appoint_time <- which.min(doc_appoint_time)
    wait_time <- max(c(0, doc_appoint_time[min_doc_appoint_time] - time_arrival))
    if (wait_time > 0) {
      # Patients waiting for a doctor
      tot_waiting_patients <- tot_waiting_patients + 1
      waiting_time <- append(waiting_time, wait_time)
    }
    # time spent by doc on each patient is uniform between 5 to 20 mins
    doc_appoint_time[min_doc_appoint_time] <- time_arrival + wait_time + runif(1, 5, 20)
  }
  # avg wait time for patients 
  if (length(waiting_time) == 0) {
    avg_waittime <- 0
  }else {
    avg_waittime <- mean(waiting_time)
  }
  
  # Office closing time
  office_close <- max(doc_appoint_time)
  
  avg_waiting_time_l <- append(avg_waiting_time_l, avg_waittime)
  office_close_time_l <- append(office_close_time_l, office_close)
  tot_patients_list <- append(tot_patients_list, tot_patients)
  tot_patients_waiting_l <- append(tot_patients_waiting_l, tot_waiting_patients)
  
}

#calculating the median values
median_avg_waiting_time<- median(avg_waiting_time_l)
median_office_close_time <- median(office_close_time_l)
median_tot_patients <- median(tot_patients_list)
median_tot_patients_waiting <- median(tot_patients_waiting_l)

office_ct <- median_office_close_time - max_time
# summary of medians
df1b <- data.frame(median_tot_patients = median_tot_patients,median_tot_patients_waiting =    median_tot_patients_waiting,median_avg_waiting_time = median_avg_waiting_time)
df1b 
print(paste(" Office close time: 4:",as.integer(office_ct),"p.m."))
```
## Section 2: Data Analysis Question 1
### 1 (a)
Sub-setting Q1Data1.dta file.


```{r}
# Section 2: Data Analysis Question 1

#___ Question 1 (a) 1)_______
library(foreign)
d1_r <- read.dta("Q1Data1.dta")

# data frame of only four columns “state,” “marital,” “heat2,” and “heat4”
dfset1 <- data.frame(state = d1_r$state , marital = d1_r$marital , heat2 = d1_r$heat2 , heat4 = d1_r$heat4, stringsAsFactors=FALSE)
head(dfset1)
nrow(dfset1)
# Sub-setting the data by removing states Hawaii, Alaska, and Washington D.C from the data frame
dfset1 <- dfset1[-c(which(dfset1$state == 'hawaii' | dfset1$state == 'alaska' | dfset1$state == 'washington dc')), ]
sum(dfset1$state == 'hawaii' | dfset1$state == 'alaska' | dfset1$state == 'washington dc')
nrow(dfset1) 

#______ Question 1 (a) 2) ______
#converting heat2 and heat4 to character
dfset1$heat2 <- as.character(dfset1$heat2)
dfset1$heat4 <- as.character(dfset1$heat4)
i_toberemoved <- c()
for(i in 1:nrow(dfset1)){
  if (is.na(dfset1[i,3])){
    #If neither of “heat2” and “heat4” has data, erase the corresponding row
    if (is.na(dfset1[i,4])){
      i_toberemoved <- append(i_toberemoved, i)
      
    #If no data is available in “heat2,” replace na for the corresponding value in “heat4”   
    }else{
      dfset1[i,3] <- dfset1[i,4]
    }
  }
}
dfset1 <- dfset1[-i_toberemoved,]
sum(is.na(dfset1$heat2))
#converting heat2 and heat4 to factor
dfset1$heat2 <- as.factor(dfset1$heat2)
dfset1$heat4 <- as.factor(dfset1$heat4)

#______ Question 1 (a) 3) _________
#3) Sub-setting the data so that it only have “dem/lean dem” and “rep/lean rep” in the “heat2” column
dfset1 <- dfset1[c(which(dfset1$heat2 == 'dem/lean dem' | dfset1$heat2 == 'rep/lean rep')), ]

#_______ Question 1 (a) 4) ________
#convert marital to character
dfset1$marital <- as.character(dfset1$marital)
imarital_na <- c()

for(i in 1:nrow(dfset1)){
  if (is.na(dfset1[i,2])){
      imarital_na <- append(imarital_na, i)
      
  }else if (dfset1[i,2] != "married"){
    # Change the label of all the variables but ‘married’ (married people) in the “marital” column to ‘other’ 
      dfset1[i,2] <- "other"
  }
}
#'na' in marital to be removed in the data frame
dfset1 <- dfset1[-imarital_na,]

#convert marital back to factor
dfset1$marital <- as.factor(dfset1$marital)
summary(dfset1$marital)
print(paste("total rows in Q1Data1 after subsetting",nrow(dfset1)))

# removing levels of removed rows
dfset1 = droplevels(dfset1)
```
### 1 (b)
Calculating proportions for each state
```{r}
states <- sort(unique(dfset1$state))
n_states <- length(states)

# Creating empty dataset of 48 rows to store values
dfresult <- data.frame(state = states, b_1 = rep(NA,n_states), b_2 = rep(NA,n_states),b_3 = rep(NA,n_states), 
                       b_4 = rep(NA,n_states), b_5 = rep(NA,n_states), actual_marriage_gap = rep(NA,n_states))

for (i in 1:nrow(dfresult)){
  
  #Multiply all values by 100 to convert to percentage
  #___ Question 1 (b) 1)____
  # the proportion of the democratic supporters
  democratic_supporters <- nrow(dfset1[dfset1$state == dfresult[i,1] & dfset1$heat2 == "dem/lean dem",])
  tot_candidates_state <- nrow(dfset1[dfset1$state==dfresult[i,1],])
  dfresult[i,2] <- (democratic_supporters/tot_candidates_state)*100
  
  #___ Question 1 (b) 2) ___
  # the proportion of the married people
  married_people <- nrow(dfset1[dfset1$state==dfresult[i,1] & dfset1$marital == "married",])
  dfresult[i,3] <- (married_people/tot_candidates_state)*100
  
  #___ Question 1 (b) 3) ____
  # the ratio of the married people among the democratic supporters to the total married people
  married_democratic <- nrow(dfset1[dfset1$state==dfresult[i,1] & dfset1$marital == "married" & dfset1$heat2 == "dem/lean dem",])
  dfresult[i,4] <- (married_democratic/married_people)*100
  
  #____ Question 1 (b) 4) ___
  #  the ratio of non-married among the democratic to the total non-married people
  nonmarried_democratic <- nrow(dfset1[dfset1$state==dfresult[i,1] & dfset1$marital == "other" & dfset1$heat2 == "dem/lean dem",])
  dfresult[i,5] <- (nonmarried_democratic/(tot_candidates_state-married_people))*100
  
  #_____ Question 1 (b) 5) ____
  # the difference of 3) and 4).
  dfresult[i,6] <- dfresult[i,4] - dfresult[i,5]
  
  # the marital gap
  dfresult[i,7] = abs(married_democratic - nonmarried_democratic)
}
# the first 5 observations of these new variables
head(dfresult,5)
```
### 1 (c)
Subsetting the data in Q1Data2.csv file
```{r}
#_____ Question 1 (c) 1) ____
# Sub-setting the data by removing states Hawaii, Alaska, and Washington D.C from the data frame
d2_r <- read.csv('Q1Data2.csv',header = T)
dfset2 <- data.frame(d2_r)
dfset2 <- dfset2[-c(which(dfset2$state == 'Hawaii' | dfset2$state == 'Alaska' | dfset2$state == 'District of Columbia')),]

#_____ Question 1 (c) 2) _____
#sub-setting only two columns “state” and “vote_Obama_pct” (Obama’s actual vote share)
dfset2 <- data.frame(state = dfset2$state , vote_Obama_pct = dfset2$vote_Obama_pct, stringsAsFactors=FALSE)

dfset2 = droplevels(dfset2)
# the first 5 lines of the data set
head(dfset2, 5)

```
### 1 (d)
 Using logistic regression for predicting vote intention
 Assumption 1: No state-level heterogeneity : Complete Pooling, 
 Assumption 2: Complete state-level heterogeneity : No Pooling, 
 Assumption 3: State-level heterogeneity is unknown : Partial Pooling
```{r}
#_____ Question 1 (d) _______
library(glmnet)
# Question 1 (d) Assumption 1 No state-level heterogeneity : Complete Pooling
# using lasso
y <- dfset1$heat2
mm_predictor1 <- model.matrix(heat2~0+marital , data=dfset1[1:3])
complete_pooling <- glmnet(mm_predictor1, y, alpha = 1, lambda = 0, family = "binomial")
coef(complete_pooling)

# Question 1 (d) Assumption 2 Complete state-level heterogeneity : No Pooling
# using lasso
mm_predictor2 <- model.matrix(heat2~0+state + marital + state*marital , data=dfset1[1:3])
no_pooling <- glmnet(mm_predictor2, y, alpha = 1, lambda = 0, family = "binomial")
coef(no_pooling)

# Question 1 (d) Assumption 3: State-level heterogeneity is unknown : 
# partially pooled on state intercepts and slopes. Outlying coefficients are penalized.
# using lasso

mm_predictor3 <- model.matrix(heat2~0+marital+state + state*marital , data=dfset1[1:3])
# setting penalty factor for marital "married" and "other" column to 0 as marital is not penalized
penaltyfactor <- rep(1,ncol(mm_predictor3))
penaltyfactor[1] <- 0
penaltyfactor[2] <- 0
# calculating best lambda value
cv_model <- cv.glmnet(mm_predictor3, y, alpha = 1,family = "binomial", penalty.factor = penaltyfactor)
best_lambda <- cv_model$lambda.min

partial_pooling <- glmnet(mm_predictor3, y, alpha = 1, lambda = best_lambda, family = "binomial", 
                          penalty.factor = penaltyfactor)
coef(partial_pooling)

```



### 1 (e) The Votes using Partial Pooling Model

Democratic voters are calculated from the partial pooling model and then it can be compared from the actual votes. In the graph the square window shows the actual democratic voters to the diamond shaped obama voters with the colorful different states inverted triangles.


```{r}
#_____ Question 1 (e) _______

# calculating prediction probability for partial pooling Assumption 1
pred_parpooling_prob <- predict(partial_pooling, s = best_lambda, newx = mm_predictor3,  type = "class")

# Creating partial pooling data frame to store values
df_partial_pooling <- data.frame(state = dfset1$state, marital = dfset1$marital, actul_op = y, pred_parpooling_prob = pred_parpooling_prob )
    
# data frame to store values of predicted vote share by state, along with actual vote intention
# and  Obama’s actual vote share for partial pooling
df_partialpool_plotting <- data.frame(state = states, predicted_parpooling_prob = rep(NA,n_states),
                          actual = dfresult$b_1, vote_Obama_pct = dfset2$vote_Obama_pct)

for (i in 1:nrow(df_partialpool_plotting)){
  
  # the proportion of the democratic supporters in partial pooling
  democratic_supporters <- nrow(df_partial_pooling[df_partial_pooling$state == df_partialpool_plotting[i,1] 
                                                   & df_partial_pooling$s1 == "dem/lean dem",])
  tot_candidates_state <- nrow(df_partial_pooling[df_partial_pooling$state==df_partialpool_plotting[i,1],])
  df_partialpool_plotting[i,2] <- (democratic_supporters/tot_candidates_state)*100
}

# plotting votes v/s states using partial pooling

library(ggplot2)
ggplot(data = df_partialpool_plotting)+
  #inverted triangle shape
  geom_point(aes(x=state,y=predicted_parpooling_prob),shape = 6,color=1:48) +
  #square window shape
  geom_point(aes(x=state,y=actual),shape = 12,color=1:48)+ 
  #diamond shape
  geom_point(aes(x=state,y=vote_Obama_pct),shape = 18,color=1:48) + 
  labs(x= "States", y = "percentage of votes",title = "Democratic Party Vote Analysis(Partial Pool)")+
  theme(axis.text.x = element_text(size = 9, angle = 90, hjust = 0.5, vjust = 0.5),
        axis.text.y = element_text(colour = "grey20", size = 12))

```



### 1 (f) The marital Gap using Partial Pooling

the predicted marital gap will be the difference between the democratic married people to democratic other people in the partial pooling model. Comparing this from the actual marital gap in the dataset we can see in the plotted graph the window shaped are the actual marital gap and the diamond shape are the obama votes for all the states represented by the inverted colorful triangles.
```{r}
#_____ Question 1 (f) _______
# The marriage gap is defined as the difference of Obama’s vote share 
# among married and non-married people (“other”) for partial pooling

df_partialpool_maritalplot <- data.frame(state = states, predicted_marriage_gap_pp = rep(NA,n_states),
                              actual_predicted_marriage_gap = dfresult$actual_marriage_gap, vote_Obama_pct = dfset2$vote_Obama_pct)

for (i in 1:nrow(df_partialpool_maritalplot)){
  
  # the number of the democratic married supporters state wise for partial pooling
  democratic_supporters_marr <- nrow(df_partial_pooling[df_partial_pooling$state == df_partialpool_maritalplot[i,1] & 
                                     df_partial_pooling$marital == "married"  & 
                                     df_partial_pooling$s1 == "dem/lean dem",])
  
  # the number of the democratic other supporters state wise for partial pooling
  democratic_supporters_oth <- nrow(df_partial_pooling[df_partial_pooling$state == df_partialpool_maritalplot[i,1] & 
                                                         df_partial_pooling$marital == "other"  & 
                                                         df_partial_pooling$s1 == "dem/lean dem",])
  df_partialpool_maritalplot[i,2] <- abs(democratic_supporters_marr - democratic_supporters_oth)
}

# plotting marital gap v/s states using partial pooling

library(ggplot2)
ggplot(data = df_partialpool_maritalplot)+
  #inverted triangle shape
  geom_point(aes(x=state,y=predicted_marriage_gap_pp),shape = 6,color=1:48) +
  #square window shape
  geom_point(aes(x=state,y=actual_predicted_marriage_gap),shape = 12,color=1:48)+ 
  #diamond shape
  geom_point(aes(x=state,y=vote_Obama_pct),shape = 18,color=1:48) + 
  ylim(0,300)+
  labs(x= "States", y = "marital gap",title = "Democratic Marital Gap Modal Analysis(Partial Pool)")+
  theme(axis.text.x = element_text(size = 9, angle = 90, hjust = 0.5, vjust = 0.5),
        axis.text.y = element_text(colour = "grey20", size = 12))

```


### 1 (g)  The Votes and The marital Gap using No Pooling

the predicted marital gap will be the difference between the democratic married people to democratic other people in the no pooling model. Comparing this from the actual marital gap in the dataset we can see in the plotted graph the window shaped are the actual marital gap and the diamond shape are the obama votes for all the states represented by the inverted colorful triangles. Democratic voters are calculated from the no pooling model and then it can be compared from the actual votes. In the graph the square window shows the actual democratic voters to the diamond shaped obama voters with the colorful different states inverted triangles.
```{r}
#_____ Question 1 (g) _______
# calculating prediction probability for no pooling Assumption 2
pred_nopooling_prob <- predict(no_pooling, s = best_lambda, newx = mm_predictor2,  type = "class")

df_no_pooling <- data.frame(state = dfset1$state, marital = dfset1$marital, actul_op = y, pred_nopooling_prob = pred_nopooling_prob )
# data frame to store values of predicted vote share by state, along with actual vote intention
# and  Obama’s actual vote share for no pooling
df_nopool_plotting <- data.frame(state = states, predicted_nopooling_prob = rep(NA,n_states),
                                      actual = dfresult$b_1, vote_Obama_pct = dfset2$vote_Obama_pct)

for (i in 1:nrow(df_nopool_plotting)){
  
  # the proportion of the democratic supporters
  democratic_supporters_np <- nrow(df_no_pooling[df_no_pooling$state == df_nopool_plotting[i,1] 
                                                   & df_no_pooling$s1 == "dem/lean dem",])
  tot_candidates_state <- nrow(df_no_pooling[df_no_pooling$state==df_nopool_plotting[i,1],])
  df_nopool_plotting[i,2] <- (democratic_supporters_np/tot_candidates_state)*100
}

# plotting votes v/s states using no pooling
ggplot(data = df_nopool_plotting)+
  #inverted triangle shape
  geom_point(aes(x=state,y=predicted_nopooling_prob),shape = 6,color=1:48) +
  #square window shape
  geom_point(aes(x=state,y=actual),shape = 12,color=1:48)+ 
  #diamond shape
  geom_point(aes(x=state,y=vote_Obama_pct),shape = 18,color=1:48) + 
  labs(x= "States", y = "percentage of votes",title = "Democratic Party Vote Analysis(No Pool)")+
  theme(axis.text.x = element_text(size = 9, angle = 90, hjust = 0.5, vjust = 0.5),
        axis.text.y = element_text(colour = "grey20", size = 12))


# The marriage gap is defined as the difference of Obama’s vote share 
# among married and non-married people (“other”) for no pooling
df_nopool_maritalplot <- data.frame(state = states, predicted_marriage_gap_np = rep(NA,n_states),
                                         actual_predicted_marriage_gap = dfresult$actual_marriage_gap, 
                                         vote_Obama_pct = dfset2$vote_Obama_pct)

for (i in 1:nrow(df_nopool_maritalplot)){
  
  # the number of the democratic married supporters state wise for no pooling
  democratic_supporters_marr_np <- nrow(df_no_pooling[df_no_pooling$state == df_nopool_maritalplot[i,1] & 
                                                          df_no_pooling$marital == "married"  & 
                                                          df_no_pooling$s1 == "dem/lean dem",])
  
  # the number of the democratic other supporters state wise  fop no pooling
  democratic_supporters_oth_np <- nrow(df_no_pooling[df_no_pooling$state == df_nopool_maritalplot[i,1] & 
                                                         df_no_pooling$marital == "other"  & 
                                                         df_no_pooling$s1 == "dem/lean dem",])
  df_nopool_maritalplot[i,2] <- abs(democratic_supporters_marr_np - democratic_supporters_oth_np)
}

# plotting marital gap v/s states using no pooling

library(ggplot2)
ggplot(data = df_nopool_maritalplot)+
  #inverted triangle shape
  geom_point(aes(x=state,y=predicted_marriage_gap_np),shape = 6,color=1:48) +
   #square window shape
  geom_point(aes(x=state,y=actual_predicted_marriage_gap),shape = 12,color=1:48)+
  #diamond shape
  geom_point(aes(x=state,y=vote_Obama_pct),shape = 18,color=1:48) + 
  ylim(0,300)+
  labs(x= "States", y = "marital gap",title = "Democratic Marital Gap Modal Analysis(No Pool)")+
  theme(axis.text.x = element_text(size = 9, angle = 90, hjust = 0.5, vjust = 0.5),
        axis.text.y = element_text(colour = "grey20", size = 12))

  

```
